# Clinical tables {#sec-omop_clinical_tables}


```{r, warning=FALSE, message=FALSE}
library(DBI)
library(dbplyr)
library(dplyr)
library(here)
library(CDMConnector)
library(ggplot2)
library(clock)
```

```{r}
db<-dbConnect(duckdb::duckdb(), 
              dbdir = eunomiaDir(datasetName = "synthea-covid19-10k"))
cdm <- cdmFromCon(db, cdmSchema = "main", writeSchema = "main", writePrefix = "my_study_")
```

```{r, echo = TRUE}
cdm
```

## Vocabulary tables

Above we've got counts by specific concept IDs recorded in the condition occurrence table. What these IDs represent is described in the concept table. Here we have the name associated with the concept, along with other information such as it's domain and vocabulary id.

```{r, echo = TRUE}
cdm$concept |> 
  glimpse()
```

Other vocabulary tables capture other information about concepts, such as the direct relationships between concepts (the concept relationship table) and hierarchical relationships between (the concept ancestor table).

```{r, echo = TRUE}
cdm$concept_relationship |> 
  glimpse()

cdm$concept_ancestor |> 
  glimpse()
```

More information on the vocabulary tables (as well as other tables in the OMOP CDM version 5.3) can be found at <https://ohdsi.github.io/CommonDataModel/cdm53.html#Vocabulary_Tables>.


